Non-word repetition in Yélî Dnye

### 2021-02-03 manuscript.Rmd
comments and minor edits through the end

## TODO Alex
- Shorten the abstract a little please (it's very clearly written now but probably too long)
- Other keywords for AC to consider adding: Papuan, non-industrial/non-urban, comparative, typology/markedness
- Changed "Scores" to "scores" in the figures—check if I missed anything!
- Check whether you're okay with the mini summary I put in on Peute's study

## TODO Middy

- read over whole thing -- does the logic sound ok?
- tell me if you think you CAN'T LIVE with my turning this in with  figures 1 and 5 as they are (I promise I'll improve them in the revision stage)

@@

Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals, as per our third research question. As shown in Figure \@ref(fig:fig-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`). In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).

@@

We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments). We aimed to provide additional data on two questions already visitied in NWR work, namely the influence of stimulus length and individual variation, and one research area that has received less attention, regarding the possible relationship between phone frequency and NWR scores. An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and other that will be discussed below). We consider implications of our results on each of these four research areas in turn.

@@

First, it remains possible that apparent length effects are actually due to uncontrolled aspects of the stimuli. For instance, some NWR researchers model their non-words on existing words, by changing some vowels and consonants -- which could lead to fewer errors (since children have produced similar words in the past); some researchers control tightly the diphone frequency of subsequences in the non-words. Building on these two aspects that researchers often control, one can imagine that longer items have fewer neighbors, and thus both the frequency with which children have produced similar items and (relatedly) their n-phone frequency is overall lower. If this idea is correct, a careful analysis of non-words used in previous work may reveal that studies with larger length effects just happened to have longer non-words with lower n-phone frequencies.

@@

To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring. We entered data from `r length(levels(pl_overall$Study))` studies (including ours), presenting data from `r length(all_lang)` languages. Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye by the present study. Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability. Results separating different age groups are shown in \@ref(fig:fig-prevlit-overall).

@@

That said, @cristia2020infant also pointed out another characteristic of the Tsimane' culture, and this was the relatively low prevalence of literacy, and generally the variable access to formal education. This is a very different case from the Yélî population studied here, where all adults have accumulated several years of schooling, and literacy is widespread. If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks. Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in \@ref(fig:fig-prevlit-overall). These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English. What is at stake in the second interpretation of the lower scores observed among the Tsimane' is related  to literacy *in the broader population* (rather than in the tested children themselves).

>>>>

## TODO Alex
- Check for my comments and minor changes throughout in <!----> format (I didn't recompile anything)
- Shorten the abstract a little please (it's very clearly written now but probably too long)
- Other keywords for AC to consider adding: Papuan, non-industrial/non-urban, comparative, typology/markedness
- Changed "Scores" to "scores" in the figures—check if I missed anything!
- Check whether you're okay with the mini summary I put in on Peute's study
- Logic sounds great! The discussion is also much improved, thank you!
- Please see my notes on the mini ethnographic info section b/c there are some inaccurate remarks right now
- I only had one small comment on figure 5
- I can live with figure 1 for now but you're right I don't love it :)

@@

Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals, as per our third research question. As shown in Figure \@ref(fig:fig-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`). In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children<!--note this ends up saying data is missing for "15" children whereas the participants section mentions it's missing for "14" (`r sum(is.na(inc$birthOrder))` in this rmd l.491) -->, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).

@@

We used non-word repetition to investigate phonological development in a language with a large phonological inventory (including some typologically rare segments). We aimed to provide additional data on two questions already visited<!--fixed sp "visited"--> in NWR work, namely the influence of stimulus length and individual variation, and one research area that has received less attention, regarding the possible relationship between phone frequency and NWR scores. An additional overarching goal was to discuss NWR in the context of population and language diversity, since it is very commonly used to document phonological development in children raised in urban settings with wide-spread literacy, and has been less seldom used in non-European languages (but note there are exceptions, including work cited in the Introduction and in the discussion below<!--minor rewording-->). We consider implications of our results on each of these four research areas in turn.

@@

First, it remains possible that apparent length effects are actually due to uncontrolled aspects of the stimuli. For instance, some NWR researchers model their non-words on existing words, by changing some vowels and consonants,<!--substituted "," for " -- "--> which could lead to fewer errors (since children have produced similar words in the past); some researchers control tightly the diphone frequency of subsequences in the non-words. Building on these two aspects that researchers often control, one can imagine that longer items have fewer neighbors, and thus both the frequency with which children have produced similar items and (relatedly) their n-phone frequency is overall lower. If this idea is correct, a careful analysis of non-words used in previous work may reveal that studies with larger length effects just happened to have longer non-words with lower n-phone frequencies.

@@

To enrich this discussion, we looked for previous studies on monolingual children with normative development learning diverse languages, and entered them when they reported non-word repetition scores based on whole item scoring. We entered data from `r length(levels(pl_overall$Study))` studies (including ours), presenting data from `r length(all_lang)` languages. Specifically, Arabic was represented by @jabere2018xperiment; Cantonese by @stokes2006nonword; English by @vance2005speech; Italian by @piazzalunga2019articulatory; Mandarin by @lei2011developmental; Persian by @farmani2018normalization; Slovak by @kapalkova2013non and @polivsenska2014improving; Sotho by @wilsenach2013phonological; Spanish by @balladares2016socio; Swedish by @kalnak2014nonword and @radeborg2006swedish; Tsimane' by @cristia2020infant; and Yélî Dnye by the present study. Studies varied in the length of non-words that were considered; whenever results were reported separately for different lengths, we calculated overall averages based on lengths of 2 and 3 syllables, for increased comparability. Results separating different age groups are shown in Figure<!--added "Figure"--> \@ref(fig:fig-prevlit-overall).<!--looks great! only let's make swedish a tinge darker gray or some kind of turqouise because it's confusable with mandarin now-->

@@

That said, @cristia2020infant also pointed out another characteristic of the Tsimane' culture, and this was the relatively low prevalence of literacy, and generally the variable access to formal education. This is a very different case from the Yélî population studied here, where nearly<!--added "nearly"--> all adults have accumulated several years of schooling, and basic<!--added "basic"--> literacy in English (and sometimes Yélî Dnye)<!--added language info--> is widespread. If this second hypothesis holds, then this may mean that there are phonetic effects of learning to read in the input afforded to young children, and that this has consequences for young children's encoding and decoding of sounds in the context of NWR tasks. Notice that this is not the same as the oft-recorded effect of learning to read affecting NWR performance, illustrated for instance in the data for Sotho in \@ref(fig:fig-prevlit-overall). These two data points have been gathered from two groups of children, all exposed mainly to Sotho, but children with higher NWR had been learning to read in Sotho, whereas those with lower scores were learning to read in English. What is at stake in the second interpretation of the lower scores observed among the Tsimane' is related  to literacy *in the broader population* (rather than in the tested children themselves).
